---
title: "Descriptive Statistics"
subtitle: "What to do with lots of numbers"
author: "Matthew Crump"
date: "2018/07/20 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["defaultb.css", "metropolisb.css", "metropolis-fontsb.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE, cache = TRUE)
```

class: pink, center, middle, clear

# What do lots of number look like?

---

# Lots of Numbers look like this

Like this

<div class=rtable>
```{r}
a<-matrix(round(runif(200,-100,100)),ncol=20)
knitr::kable(a, 'html')
```
</div>

---


# What can we say about them?

We can see they aren't all the same. Not much else really. Looking at a bunch of numbers is hard work.

<div class=rtable>
```{r}
knitr::kable(a, 'html')
```
</div>

---


# Summary numbers

It would be nice to reduce the big set of numbers down to a few numbers that we can look at and make sense of.

## Sameness (Central Tendency)

- What are all the numbers close to?

## Differentness (Variance)

- How different are the numbers?

---


# Descriptive Statistics 

- Give us summaries of big sets of numbers

- Useful single numbers to look at

- They tell us about patterns of sameness and differentness


---


class: pink, center, middle, clear

# Graph the numbers to get a better look

---


# Dot plot (unordered)

Graphing the numbers gives a quick and dirty sense of what they are like. Here's 200 numbers presented as dots

```{r, fig.align='center', fig.height=6}
library(ggplot2)
qplot(y=rnorm(200,0,1))+
  geom_point()+
  theme_classic(base_size = 30)+
  ylab("value")
```

---



# Dot plot (ordered)

Sorting the numbers from smallest to largest

```{r, fig.align='center'}
a<-sort(rnorm(200,0,1))
qplot(y=a)+
  geom_point()+
  theme_classic(base_size = 30)+
  ylab("value")
```

---



# Histograms

Histograms count up the numbers inside specific ranges

```{r, fig.align='center', fig.height=5}
qplot(x=a)+
  geom_histogram(color="white")+
  theme_classic(base_size = 30)+
  ylab("count")+
  xlab("value")
```

---

# Histograms

Bars show you which bins have more or less numbers in the range

```{r, fig.align='center', fig.height=5}
qplot(x=a)+
  geom_histogram(color="white")+
  theme_classic(base_size = 30)+
  ylab("count")+
  xlab("value")
```


---



# So what are these numbers like?

What single number would you say best describes most of these numbers?

```{r, fig.align='center', fig.height=6}
qplot(x=a)+
  geom_histogram(color="white")+
  theme_classic(base_size = 30)+
  ylab("count")+
  xlab("value")
```


---




# Question

Is the red or blue value a better summary of all the numbers?

```{r, fig.align='center', fig.height=6}
qplot(x=a)+
  geom_histogram(color="white")+
  theme_classic(base_size = 30)+
  ylab("count")+
  geom_vline(xintercept=mean(a),size=3, color="red")+
  geom_vline(xintercept=mean(a)-2,size=3, color="blue")+
  xlab("value")
```

---


class: pink, center, middle, clear

# Measures of Central Tendency

---

# Central Tendency

1. **Central tendency** should describe what most of the data is like

--

2. We want our summary number to be most like the other numbers. We want it to be a **representative value**

--

3. There are **multiple measures** of central tendency

--

4. They have **different properties**

--

5. Some work better than others depending on the data

---



# Mode

The mode is the single most frequently occuring number 

> 1 1 2 2 3 4 5 6 7 7 7 7 7

- The mode is 7 because 7 happens the most

- Find the mode by counting the occurence of each number, the mode is the most frequently occuring number

- If there is a tie, then you have two or three or more modes (depends on how many differen numbers tie)

---




# Finding the Mode in R

We make 25 numbers, how do we get R to find the mode?

```{r, echo=TRUE, eval=TRUE}
#make some numbers
a <- round(rnorm(n=25, mean=24, sd=5)) 
```


```{r, class.output='font50', cache=TRUE, eval=TRUE}
knitr::kable(matrix(a,ncol=5))
```

---




# Finding the Mode in R

`table` function automatically counts the occurence of each number

```{r, echo=TRUE, results='markup'}
table(a)
```

We can see that 29 occurs the most

---


# Custom function for the mode in R

You can always write your own function for the mode. This one is called `my_mode`

```{r, echo=TRUE}
my_mode <- function(x) {
     ux <- unique(x)
     ux[which.max(tabulate(match(x, ux)))]
 }
```

```{r, echo=TRUE, cache=FALSE}
my_mode(a)
```

---

# Thinking about the mode

1. Tells us the most frequent number(s)

--

2. Is it representative of all the numbers?

--

3. When would the mode be a good thing to know?

---



