# Estimating where your data came from: Of samples and populations

<span class="newthought">
I have studied many languages-French, Spanish and a little Italian, but no one told me that Statistics was a foreign language.
---Charmaine J. Forde
</span>

<div class="marginnote">
Original text by Danielle Navarro
Some edits and additions by Matthew Crump
</div>

Remember, the role of descriptive statistics is to concisely summarise what we **do** know. In contrast, the purpose of inferential statistics is to "learn what we do not know from what we do". What kinds of things would we like to learn about? And how do we learn them? These are the questions that lie at the heart of inferential statistics, and they are traditionally divided into two "big ideas": estimation and hypothesis testing. The goal in this chapter is to introduce the first of these big ideas, estimation theory, but we'll talk about sampling theory first because estimation theory doesn't make sense until you understand sampling. So, this chapter divides into sampling theory, and how to make use of sampling theory to discuss how statisticians think about estimation. We have already done lots of sampling, so you are already familiar with some of the big ideas.

## Samples, populations and sampling

**Sampling theory** plays a huge role in specifying the assumptions upon which your statistical inferences rely. And in order to talk about "making inferences" the way statisticians think about it, we need to be a bit more explicit about what it is that we're drawing inferences **from** (the sample) and what it is that we're drawing inferences **about** (the population).

In almost every situation of interest, what we have available to us as researchers is a **sample** of data. We might have run experiment with some number of participants; a polling company might have phoned some number of people to ask questions about voting intentions; etc. Regardless: the data set available to us is finite, and incomplete. We can't possibly get every person in the world to do our experiment; a polling company doesn't have the time or the money to ring up every voter in the country etc. In our earlier discussion of descriptive statistics, this sample was the only thing we were interested in. Our only goal was to find ways of describing, summarising and graphing that sample. This is about to change.

### Defining a population

A sample is a concrete thing. You can open up a data file, and there's the data from your sample. A **population**, on the other hand, is a more abstract idea. It refers to the set of all possible people, or all possible observations, that you want to draw conclusions about, and is generally **much** bigger than the sample. In an ideal world, the researcher would begin the study with a clear idea of what the population of interest is, since the process of designing a study and testing hypotheses about the data that it produces does depend on the population about which you want to make statements. However, that doesn't always happen in practice: usually the researcher has a fairly vague idea of what the population is and designs the study as best he/she can on that basis.

Sometimes it's easy to state the population of interest. For instance, in the "polling company" example, the population consisted of all voters enrolled at the a time of the study – millions of people. The sample was a set of 1000 people who all belong to that population. In most situations the situation is much less simple. In a typical a psychological experiment, determining the population of interest is a bit more complicated. Suppose I run an experiment using 100 undergraduate students as my participants. My goal, as a cognitive scientist, is to try to learn something about how the mind works. So, which of the following would count as "the population":

-   All of the undergraduate psychology students at the University of Adelaide?

-   Undergraduate psychology students in general, anywhere in the world?

-   Australians currently living?

-   Australians of similar ages to my sample?

-   Anyone currently alive?

-   Any human being, past, present or future?

-   Any biological organism with a sufficient degree of intelligence operating in a terrestrial environment?

-   Any intelligent being?

Each of these defines a real group of mind-possessing entities, all of which might be of interest to me as a cognitive scientist, and it's not at all clear which one ought to be the true population of interest. 

### Simple random samples

Irrespective of how we define the population, the critical point is that the sample is a subset of the population, and our goal is to use our knowledge of the sample to draw inferences about the properties of the population. The relationship between the two depends on the **procedure** by which the sample was selected. This procedure is referred to as a **sampling method**, and it is important to understand why it matters.

To keep things simple, imagine we have a bag containing 10 chips. Each chip has a unique letter printed on it, so we can distinguish between the 10 chips. The chips come in two colors, black and white. 

```{r srs1, fig.cap="Simple random sampling without replacement from a finite population", echo=FALSE,dev='png'}
knitr::include_graphics("navarro_img/estimation/srs1.pdf")
```

This set of chips is the population of interest, and it is depicted graphically on the left of Figure \@ref(fig:srs1). 

As you can see from looking at the picture, there are 4 black chips and 6 white chips, but of course in real life we wouldn't know that unless we looked in the bag. Now imagine you run the following "experiment": you shake up the bag, close your eyes, and pull out 4 chips without putting any of them back into the bag. First out comes the $a$ chip (black), then the $c$ chip (white), then $j$ (white) and then finally $b$ (black). If you wanted, you could then put all the chips back in the bag and repeat the experiment, as depicted on the right hand side of Figure\@ref(fig:srs1). Each time you get different results, but the procedure is identical in each case. The fact that the same procedure can lead to different results each time, we refer to it as a **random** process. However, because we shook the bag before pulling any chips out, it seems reasonable to think that every chip has the same chance of being selected. A procedure in which every member of the population has the same chance of being selected is called a **simple random sample**. The fact that we did **not** put the chips back in the bag after pulling them out means that you can't observe the same thing twice, and in such cases the observations are said to have been sampled **without replacement**.

To help make sure you understand the importance of the sampling procedure, consider an alternative way in which the experiment could have been run. Suppose that my 5-year old son had opened the bag, and decided to pull out four black chips without putting any of them back in the bag. This **biased** sampling scheme is depicted in Figure \@ref(fig:brs). 

```{r brs, fig.cap="Biased sampling without replacement from a finite population", echo=FALSE,dev='png'}
knitr::include_graphics("navarro_img/estimation/brs.pdf")
```

Now consider the evidentiary value of seeing 4 black chips and 0 white chips. Clearly, it depends on the sampling scheme, does it not? If you know that the sampling scheme is biased to select only black chips, then a sample that consists of only black chips doesn't tell you very much about the population! For this reason, statisticians really like it when a data set can be considered a simple random sample, because it makes the data analysis **much** easier.

A third procedure is worth mentioning. This time around we close our eyes, shake the bag, and pull out a chip. This time, however, we record the observation and then put the chip back in the bag. Again we close our eyes, shake the bag, and pull out a chip. We then repeat this procedure until we have 4 chips. Data sets generated in this way are still simple random samples, but because we put the chips back in the bag immediately after drawing them it is referred to as a sample **with replacement**. The difference between this situation and the first one is that it is possible to observe the same population member multiple times, as illustrated in Figure \@ref(fig:srs2).

```{r srs2, fig.cap="Simple random sampling with replacement from a finite population", echo=FALSE,dev='png'}
knitr::include_graphics("navarro_img/estimation/srs2.pdf")
```

Most psychology experiments tend to be sampling without replacement, because the same person is not allowed to participate in the experiment twice. However, most statistical theory is based on the assumption that the data arise from a simple random sample **with** replacement. In real life, this very rarely matters. If the population of interest is large (e.g., has more than 10 entities!) the difference between sampling with- and without- replacement is too small to be concerned with. The difference between simple random samples and biased samples, on the other hand, is not such an easy thing to dismiss.

### Most samples are not simple random samples

As you can see from looking at the list of possible populations that I showed above, it is almost impossible to obtain a simple random sample from most populations of interest. When I run experiments, I'd consider it a minor miracle if my participants turned out to be a random sampling of the undergraduate psychology students at Adelaide university, even though this is by far the narrowest population that I might want to generalise to. A thorough discussion of other types of sampling schemes is beyond the scope of this book, but to give you a sense of what's out there I'll list a few of the more important ones:

-   **Stratified sampling**. Suppose your population is (or can be) divided into several different subpopulations, or **strata**. Perhaps you're running a study at several different sites, for example. Instead of trying to sample randomly from the population as a whole, you instead try to collect a separate random sample from each of the strata. Stratified sampling is sometimes easier to do than simple random sampling, especially when the population is already divided into the distinct strata. It can also be more efficient that simple random sampling, especially when some of the subpopulations are rare. For instance, when studying schizophrenia it would be much better to divide the population into two strata (schizophrenic and not-schizophrenic), and then sample an equal number of people from each group. If you selected people randomly, you would get so few schizophrenic people in the sample that your study would be useless. This specific kind of of stratified sampling is referred to as **oversampling** because it makes a deliberate attempt to over-represent rare groups.

-   **Snowball sampling** is a technique that is especially useful when sampling from a "hidden" or hard to access population, and is especially common in social sciences. For instance, suppose the researchers want to conduct an opinion poll among transgender people. The research team might only have contact details for a few trans folks, so the survey starts by asking them to participate (stage 1). At the end of the survey, the participants are asked to provide contact details for other people who might want to participate. In stage 2, those new contacts are surveyed. The process continues until the researchers have sufficient data. The big advantage to snowball sampling is that it gets you data in situations that might otherwise be impossible to get any. On the statistical side, the main disadvantage is that the sample is highly non-random, and non-random in ways that are difficult to address. On the real life side, the disadvantage is that the procedure can be unethical if not handled well, because hidden populations are often hidden for a reason. I chose transgender people as an example here to highlight this: if you weren't careful you might end up outing people who don't want to be outed (very, very bad form), and even if you don't make that mistake it can still be intrusive to use people's social networks to study them. It's certainly very hard to get people's informed consent **before** contacting them, yet in many cases the simple act of contacting them and saying "hey we want to study you" can be hurtful. Social networks are complex things, and just because you can use them to get data doesn't always mean you should.

-   **Convenience sampling** is more or less what it sounds like. The samples are chosen in a way that is convenient to the researcher, and not selected at random from the population of interest. Snowball sampling is one type of convenience sampling, but there are many others. A common example in psychology are studies that rely on undergraduate psychology students. These samples are generally non-random in two respects: firstly, reliance on undergraduate psychology students automatically means that your data are restricted to a single subpopulation. Secondly, the students usually get to pick which studies they participate in, so the sample is a self selected subset of psychology students not a randomly selected subset. In real life, most studies are convenience samples of one form or another. This is sometimes a severe limitation, but not always.

### How much does it matter if you don't have a simple random sample?

Okay, so real world data collection tends not to involve nice simple random samples. Does that matter? A little thought should make it clear to you that it **can** matter if your data are not a simple random sample: just think about the difference between Figures\@ref(fig:srs1) and \@ref(fig:brs). However, it's not quite as bad as it sounds. Some types of biased samples are entirely unproblematic. For instance, when using a stratified sampling technique you actually **know** what the bias is because you created it deliberately, often to **increase** the effectiveness of your study, and there are statistical techniques that you can use to adjust for the biases you've introduced (not covered in this book!). So in those situations it's not a problem.

More generally though, it's important to remember that random sampling is a means to an end, not the end in itself. Let's assume you've relied on a convenience sample, and as such you can assume it's biased. A bias in your sampling method is only a problem if it causes you to draw the wrong conclusions. When viewed from that perspective, I'd argue that we don't need the sample to be randomly generated in **every** respect: we only need it to be random with respect to the psychologically-relevant phenomenon of interest. Suppose I'm doing a study looking at working memory capacity. In study 1, I actually have the ability to sample randomly from all human beings currently alive, with one exception: I can only sample people born on a Monday. In study 2, I am able to sample randomly from the Australian population. I want to generalise my results to the population of all living humans. Which study is better? The answer, obviously, is study 1. Why? Because we have no reason to think that being "born on a Monday" has any interesting relationship to working memory capacity. In contrast, I can think of several reasons why "being Australian" might matter. Australia is a wealthy, industrialised country with a very well-developed education system. People growing up in that system will have had life experiences much more similar to the experiences of the people who designed the tests for working memory capacity. This shared experience might easily translate into similar beliefs about how to "take a test", a shared assumption about how psychological experimentation works, and so on. These things might actually matter. For instance, "test taking" style might have taught the Australian participants how to direct their attention exclusively on fairly abstract test materials relative to people that haven't grown up in a similar environment; leading to a misleading picture of what working memory capacity is.

There are two points hidden in this discussion. Firstly, when designing your own studies, it's important to think about what population you care about, and try hard to sample in a way that is appropriate to that population. In practice, you're usually forced to put up with a "sample of convenience" (e.g., psychology lecturers sample psychology students because that's the least expensive way to collect data, and our coffers aren't exactly overflowing with gold), but if so you should at least spend some time thinking about what the dangers of this practice might be.

Secondly, if you're going to criticise someone else's study because they've used a sample of convenience rather than laboriously sampling randomly from the entire human population, at least have the courtesy to offer a specific theory as to **how** this might have distorted the results. Remember, everyone in science is aware of this issue, and does what they can to alleviate it. Merely pointing out that "the study only included people from group BLAH" is entirely unhelpful, and borders on being insulting to the researchers, who are aware of the issue. They just don't happen to be in possession of the infinite supply of time and money required to construct the perfect sample. In short, if you want to offer a responsible critique of the sampling process, then be **helpful**. Rehashing the blindingly obvious truisms that I've been rambling on about in this section isn't helpful.

### Population parameters and sample statistics

Okay. Setting aside the thorny methodological issues associated with obtaining a random sample, let's consider a slightly different issue. Up to this point we have been talking about populations the way a scientist might. To a psychologist, a population might be a group of people. To an ecologist, a population might be a group of bears. In most cases the populations that scientists care about are concrete things that actually exist in the real world. 

Statisticians, however, are a funny lot. On the one hand, they **are** interested in real world data and real science in the same way that scientists are. On the other hand, they also operate in the realm of pure abstraction in the way that mathematicians do. As a consequence, statistical theory tends to be a bit abstract in how a population is defined. In much the same way that psychological researchers operationalise our abstract theoretical ideas in terms of concrete measurements, statisticians operationalise the concept of a "population" in terms of mathematical objects that they know how to work with. You've already come across these objects they're called probability distributions (remember, the place where data comes from).

The idea is quite simple. Let's say we're talking about IQ scores. To a psychologist, the population of interest is a group of actual humans who have IQ scores. A statistician "simplifies" this by operationally defining the population as the probability distribution depicted in Figure \@ref(fig:IQdist)a. 

.

```{r IQdist, fig.cap="The population distribution of IQ scores (panel a) and two samples drawn randomly from it. In panel b we have a sample of 100 observations, and panel c we have a sample of 10,000 observations.", echo=FALSE,dev='png'}
knitr::include_graphics("navarro_img/estimation/IQdist.pdf")
```

IQ tests are designed so that the average IQ is 100, the standard deviation of IQ scores is 15, and the distribution of IQ scores is normal. These values are referred to as the **population parameters** because they are characteristics of the entire population. That is, we say that the population mean $\mu$ is 100, and the population standard deviation $\sigma$ is 15.

Now suppose we collect some data. We select 100 people at random and administer an IQ test, giving a simple random sample from the population. The sample would consist of a collection of numbers like this:

`106 101 98 80 74 ... 107 72 100`

Each of these IQ scores is sampled from a normal distribution with mean 100 and standard deviation 15. So if I plot a histogram of the sample, I get something like the one shown in Figure \@ref(fig:IQdist)b. As you can see, the histogram is **roughly** the right shape, but it's a very crude approximation to the true population distribution shown in Figure \@ref(fig:IQdist)a. The mean of the sample is fairly close to the population mean 100 but not identical. In this case, it turns out that the people in the sample have a mean IQ of 98.5, and the standard deviation of their IQ scores is 15.9. These **sample statistics** are properties of the data set, and although they are fairly similar to the true population values, they are not the same. **In general, sample statistics are the things you can calculate from your data set, and the population parameters are the things you want to learn about.** Later on in this chapter we'll talk about how you can estimate population parameters using your sample statistics and how to work out how confident you are in your estimates but before we get to that there's a few more ideas in sampling theory that you need to know about.

## The law of large numbers 

We just looked at the results of one fictitious IQ experiment with a sample size of $N=100$. The results were somewhat encouraging: the true population mean is 100, and the sample mean of 98.5 is a pretty reasonable approximation to it. In many scientific studies that level of precision is perfectly acceptable, but in other situations you need to be a lot more precise. If we want our sample statistics to be much closer to the population parameters, what can we do about it?

The obvious answer is to collect more data. Suppose that we ran a much larger experiment, this time measuring the IQs of 10,000 people. We can simulate the results of this experiment using R, using the **rnorm()** function, which generates random numbers sampled from a normal distribution. For an experiment with a sample size of **n = 10000**, and a population with **mean = 100** and **sd = 15**, R produces our fake IQ data using these commands:

```{r, echo=T}
IQ <- rnorm(n=10000, mean=100, sd=15) #generate IQ scores
IQ <- round(IQ) # make round numbers
```

Cool, we just generated 10,000 fake IQ scores. Where did they go? Well, they went into the variable IQ on my computer. You can do the same on your computer too by copying the above code. 10,000 numbers is too many numbers to look at. We can look at the first 100 like this:

```{r, echo=T}
print(IQ[1:100])
```


We can compute the mean IQ using the command **mean(IQ)** and the standard deviation using the command **sd(IQ)**, and draw a histgram using **hist()**. The histogram of this much larger sample is shown in Figure \@ref(fig:IQdist)c. Even a moment's inspections makes clear that the larger sample is a much better approximation to the true population distribution than the smaller one. This is reflected in the sample statistics: the mean IQ for the larger sample turns out to be 99.9, and the standard deviation is 15.1. These values are now very close to the true population.

I feel a bit silly saying this, but the thing I want you to take away from this is that large samples generally give you better information. I feel silly saying it because it's so bloody obvious that it shouldn't need to be said. In fact, it's such an obvious point that when Jacob Bernoulli – one of the founders of probability theory – formalised this idea back in 1713, he was kind of a jerk about it. Here's how he described the fact that we all share this intuition:

> **For even the most stupid of men, by some instinct of nature, by himself and without any instruction (which is a remarkable thing), is convinced that the more observations have been made, the less danger there is of wandering from one's goal** (see Stigler, 1986, p65).

Okay, so the passage comes across as a bit condescending (not to mention sexist), but his main point is correct: it really does feel obvious that more data will give you better answers. The question is, why is this so? Not surprisingly, this intuition that we all share turns out to be correct, and statisticians refer to it as the **law of large numbers**. The law of large numbers is a mathematical law that applies to many different sample statistics, but the simplest way to think about it is as a law about averages. The sample mean is the most obvious example of a statistic that relies on averaging (because that's what the mean is... an average), so let's look at that. **When applied to the sample mean, what the law of large numbers states is that as the sample gets larger, the sample mean tends to get closer to the true population mean.** Or, to say it a little bit more precisely, as the sample size "approaches" infinity (written as $N \rightarrow \infty$) the sample mean approaches the population mean ($\bar{X} \rightarrow \mu$).

I don't intend to subject you to a proof that the law of large numbers is true, but it's one of the most important tools for statistical theory. The law of large numbers is the thing we can use to justify our belief that collecting more and more data will eventually lead us to the truth. For any particular data set, the sample statistics that we calculate from it will be wrong, but the law of large numbers tells us that if we keep collecting more data those sample statistics will tend to get closer and closer to the true population parameters.

## Sampling distributions and the central limit theorem

The law of large numbers is a very powerful tool, but it's not going to be good enough to answer all our questions. Among other things, all it gives us is a "long run guarantee". In the long run, if we were somehow able to collect an infinite amount of data, then the law of large numbers guarantees that our sample statistics will be correct. But as John Maynard Keynes famously argued in economics, a long run guarantee is of little use in real life:

> **[The] long run is a misleading guide to current affairs. In the long run we are all dead. Economists set themselves too easy, too useless a task, if in tempestuous seasons they can only tell us, that when the storm is long past, the ocean is flat again.** @Keynes1923 [p. 80]

As in economics, so too in psychology and statistics. It is not enough to know that we will **eventually** arrive at the right answer when calculating the sample mean. Knowing that an infinitely large data set will tell me the exact value of the population mean is cold comfort when my **actual** data set has a sample size of $N=100$. In real life, then, we must know something about the behaviour of the sample mean when it is calculated from a more modest data set!

### Sampling distribution of the sample means

"Oh no, what is the sample distribution of the sample means? Is that even allowed in English?". Yes, unfortunately, this is allowed. The **sampling distribution of the sample means** is the next most important thing you will need to understand. IT IS SO IMPORTANT THAT IT IS NECESSARY TO USE ALL CAPS. It is only confusing at first because it's long and uses sampling and sample in the same phrase.

Don't worry, we've been prepping you for this. You know what a distribution is right? It's where numbers comes from. It makes some numbers occur more or less frequently, or the same as other numbers. You know what a sample is right? It's the numbers we take from a distribution. So, what could the sampling distribution of the sample means refer to? 

First, what do you think the sample means refers to? Well, if you took a sample of numbers, you would have a bunch of numbers...then, you could compute the mean of those numbers. The sample mean is the mean of the numbers in the sample. That is all. So, what is this distribution you speak of? Well, what if you took a bunch of samples of, put one here, put one there, put some other ones other places. You have a lot of different samples of numbers. You could compute the mean for each them. Then you would have a bunch of means. What do those means look like? Well, if you put them in a histogram, you could find out. If you did that, you would be looking at (roughly) a distribution, AKA **the sampling distribution of the sample means**.

"I'm following along sort of, why would I want to do this instead of watching netflix...". Because, the sampling distribution of the sample means gives you another window into chance. A very useful one that you can control, just like your remote control, by pressing the right design buttons.

#### Seeing the pieces

To make a sampling distribution of the sample means, we just need the following:

1. A distribution to take numbers from
2. A bunch of different samples from the distribution
3. The means of each of the samples
4. Get all of the sample means, and plot them in a histogram

---
Question for yourself: What do you think the sampling distribution of the sample means will look like? Will it tend to look the shape of the distribution that the samples came from? Or not? Good question, think about it.
--

Let's do those four things. We already know about the uniform distribution. So, let's use that for fun. Remember, it looks like this if we are sampling from the set of integers from 1 to 10:

```{r,echo=F}
library(ggplot2)

df<-data.frame(a=1:10,b=seq(.1,1,.1))
df$a<-as.factor(df$a)

ggplot(df,aes(x=a,y=b))+
  geom_point(color="white")+
  geom_hline(yintercept=.1)+
  theme_classic()+
  ylab("Probability")+
  xlab("Number")+
  ggtitle("Uniform distribution for numbers 1 to 10")


```

Ok, now let's take a bunch of samples from that distribution. We will set our sample-size to 20. Here is a histogram of 10 of from that bunch:

```{r, echo=F}
a<-round(runif(20*10,1,10))
df<-data.frame(a,sample=rep(1:10,each=20))
df2<-aggregate(sample~a,df,mean)

ggplot() +
  geom_histogram(data = df, aes(x = sample, y = ..density..)) +
  geom_vline(data = df2, mapping = aes(xintercept = sample), color="red") +
  facet_wrap(~a)+
  scale_x_continuous(breaks=seq(1,10,1))+
  theme_classic()

```

Ok, what have we got here? We have 10 different samples. Each sample has 20 observations and these are summarized in each of the 10 histograms. Each histogram has a red line. The red line shows you where the mean of each sample is located. So, we have found the sample means for the 10 different samples from a uniform distribution. 

First question. Are the sample means all the same? The answer is no. They are all kind of similar to each other though, they are all around five plus or minus a few numbers. This is interesting. Although all of our samples look pretty different from one another, the means of our samples look more similar than different.

Second question. What should we do with the means of our samples? Well, how about we collect them them all, and then plot a histogram of them. This would allow us to see what the distribution of the sample means looks like. The next histogram is just this. Except, rather than  taking 10 samples, we will take 10,000 samples. For each of them we will compute the means. So, we will have 10,000 means. This is the histogram of the sample means:

```{r, echo=F}
a<-round(runif(20*10000,1,10))
df<-data.frame(a,sample=rep(1:10000,each=20))
df2<-aggregate(a~sample,df,mean)

ggplot(df2, aes(x=a))+
  geom_histogram(color="white")+
  theme_classic()+
  ggtitle("Histogram of 10,000 sample means")+
  xlab("value")


```

"Wait what? This doesn't look right. I thought we were taking samples from a uniform distribution. Uniform distributions are flat. I thought you said that when you take lots and lots of sample, which is what you did, that the more you have, the more they will look the distribution they came from...THIS DOES NOT LOOK LIKE A FLAT DISTRIBTUION, WHAT IS GOING ON, AAAAAGGGHH.". We feel your pain.

Remember, we are looking at the distribution of sample means. It is indeed true that the distribution of sample means does not look the same as the distribution we took the samples from. Our distribution of sample means goes up and down. In fact, this will almost always be the case for distributions of sample means. This fact is called the **central limit theorem**, which we talk about later.

For now, let's talk about about what's happening. Remember, we have been sampling numbers between the range 1 to 10. We are supposed to get each number with roughly equal frequency, because we are sampling from a uniform distribution. So, let's say we took a sample of 10 numbers, and happened to get one of each from 1 to 10. 

`1 2 3 4 5 6 7 8 9 10`

What is the mean of those numbers? Well, its 1+2+3+4+5+6+7+8+9+10 = 55 / 10 = 5.5. Imagine if we took a bigger sample, say of 20 numbers, and again we got exactly 2 of each number. What would the mean be? It would be (1+2+3+4+5+6+7+8+9+10)*2 = 110 / 20 = 5.5. Still 5.5. You can see here, that the mean value of our uniform distribution is 5.5. Now that we know this, we might expect that most of our samples will have a mean near this number. We already know that every sample won'y be perfect, and it won't have exactly an equal amount of every number. So, we will expect the mean of our samples to vary a little bit. The histogram that we made shows the variation. Not surprisingly, the numbes vary around the value 5.5. 

We can watch a movie to see how the sample mean behaves across many different samples. It's the same as what we already saw but more fun. Each histogram shows a new sample. The red line shows where the mean of the sample is. The samples are all very different from each other, but the red line doesn't move around very much, it always stays near the middle. However, the red line does move around a little bit, and this variance is what we call the sampling distribution of the sample mean.


```{r, echo=F}
a<-round(runif(20*10,1,10))
df<-data.frame(a,sample=rep(1:10,each=20))
df2<-aggregate(sample~a,df,mean)
df<-cbind(df,mean_loc=rep(df2$sample,each=20))

library(gganimate)

ggplot(df,aes(x=a, group=sample,frame=sample)) +
  geom_histogram() +
  geom_vline(aes(xintercept=mean_loc,frame = sample),color="red")+
  scale_x_continuous(breaks=seq(1,10,1))+
  theme_classic()+
  transition_states(
    sample,
    transition_length = 2,
    state_length = 1
  )+enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')

```

### Sampling distributions exist for any sample statistic!

One thing to keep in mind when thinking about sampling distributions is that **any** sample statistic you might care to calculate has a sampling distribution. For example, suppose that each time you sampled some numbers experiment you wrote down the largest number in the experiment. Doing this over and over again would give you a very different sampling distribution, namely the **sampling distribution of the maximum**. You could calculate the smallest number, or the mode, or the median, of the variance, or the standard deviation, or anything else from your sample. Then, you could repeat many times, and produce the sampling distribution of those statistics. Neat!

Just for fun here are some different sampling distributions for different statistics. We will take a normal distribution with mean = 100, and standard deviation =20. Then, we'll take lots of sample with n = 50 (50 observations per sample). We'l save all of the sample statistics, then plot their histograms. Let's do it:

```{r}
all_df<-data.frame()
for(i in 1:10000){
  sample<-rnorm(50,100,20)
  sample_mean<-mean(sample)
  sample_sd<-sd(sample)
  sample_max<-max(sample)
  sample_median<-median(sample)
  t_df<-data.frame(i,sample_mean,sample_sd,sample_max,sample_median)
  all_df<-rbind(all_df,t_df)
}

library(ggpubr)
a<-ggplot(all_df,aes(x=sample_mean))+
  geom_histogram(color="white")+
  theme_classic()
b<-ggplot(all_df,aes(x=sample_sd))+
  geom_histogram(color="white")+
  theme_classic()
c<-ggplot(all_df,aes(x=sample_max))+
  geom_histogram(color="white")+
  theme_classic()
d<-ggplot(all_df,aes(x=sample_median))+
  geom_histogram(color="white")+
  theme_classic()

ggarrange(a,b,c,d,
          ncol = 2, nrow = 2)

```

We just computed 4 different sampling distributions, for the mean, standard deviation, maximum value, and the median. If you just look quickly at these histograms you might think they all basically look the same. Hold up now. It's very important to look at the x-axes. They are different. For example, the sample mean goes from about 90 to 110, whereas the standard deviation goes from 15 to 25. Finally, while it is true that these histograms all look like normal distributions, notice that the distribution for the sample max is kind of skewed, not totally symmetrical.

These sampling distributions are super important, and worth thinking about. What should you think about? Well, here's a clue. These distributions are telling you what to expect from your sample. Critically, they are telling you what you should expect from a sample, when you take one from the specific distribution that we used (normal distribution with mean =100 and sd = 20). What have we learned. We've leard a tonne. We've learned that we can expect our sample to have a mean somewhere between 90 and 108ish. Notice, the sample means are never more extreme. We've learned that our sample with usually have some variance, and that the the standard deviation will be somewhere between 15 and 25 (never much more extreme than that). We can see that sometime we get some big numbers, say between 120 and 180, but not much bigger than tha. And, we can see that the median is pretty similar to the mean. If you ever took a sample of 50 numbers, and your descriptive statistics were inside these windows, then perhaps they came from this kind of normal distribution. If your sample statistics are very different, then your sample probably did not come this distribution. By using simulation, we can find out what samples look like when they come from distributions, and we can use this information to make inferences about whether our sample came from particular distributions.

### The central limit theorem

At this point I hope you have a pretty good sense of what sampling distributions are, and in particular what the sampling distribution of the mean is. In this section I want to talk about how the sampling distribution of the mean changes as a function of sample size. Intuitively, you already know part of the answer: if you only have a few observations, the sample mean is likely to be quite inaccurate: if you replicate a small experiment and recalculate the mean you'll get a very different answer. In other words, the sampling distribution is quite wide. If you replicate a large experiment and recalculate the sample mean you'll probably get the same answer you got last time, so the sampling distribution will be very narrow. You can see this visually in Figure [fig:IQsamp]: the bigger the sample size, the narrower the sampling distribution gets. We can quantify this effect by calculating the standard deviation of the sampling distribution, which is referred to as the **standard error**. The standard error of a statistic is often denoted SE, and since we're usually interested in the standard error of the sample **mean**, we often use the acronym SEM. As you can see just by looking at the picture, as the sample size $N$ increases, the SEM decreases.

Okay, so that's one part of the story. However, there's something I've been glossing over so far. All my examples up to this point have been based on the "IQ scores" experiments, and because IQ scores are roughly normally distributed, I've assumed that the population distribution is normal. What if it isn't normal? What happens to the sampling distribution of the mean? The remarkable thing is this: no matter what shape your population distribution is, as $N$ increases the sampling distribution of the mean starts to look more like a normal distribution. To give you a sense of this, I ran some simulations using . To do this, I started with the "ramped" distribution shown in the histogram in Figure [fig:cltdemo]. As you can see by comparing the triangular shaped histogram to the bell curve plotted by the black line, the population distribution doesn't look very much like a normal distribution at all. Next, I used  to simulate the results of a large number of experiments. In each experiment I took $N=2$ samples from this distribution, and then calculated the sample mean. Figure [fig:cltdemo]b plots the histogram of these sample means (i.e., the sampling distribution of the mean for $N=2$). This time, the histogram produces a $\cap$-shaped distribution: it's still not normal, but it's a lot closer to the black line than the population distribution in Figure [fig:cltdemo]a. When I increase the sample size to $N=4$, the sampling distribution of the mean is very close to normal (Figure [fig:cltdemo]c), and by the time we reach a sample size of $N=8$ it's almost perfectly normal. In other words, as long as your sample size isn't tiny, the sampling distribution of the mean will be approximately normal no matter what your population distribution looks like!

  ----- -----
        
   (a)   (b)
        
   (c)   (d)
  ----- -----

[fig:cltdemo]

On the basis of these figures, it seems like we have evidence for all of the following claims about the sampling distribution of the mean:

-   The mean of the sampling distribution is the same as the mean of the population

-   The standard deviation of the sampling distribution (i.e., the standard error) gets smaller as the sample size increases

-   The shape of the sampling distribution becomes normal as the sample size increases

As it happens, not only are all of these statements true, there is a very famous theorem in statistics that proves all three of them, known as the **central limit theorem**. Among other things, the central limit theorem tells us that if the population distribution has mean $\mu$ and standard deviation $\sigma$, then the sampling distribution of the mean also has mean $\mu$, and the standard error of the mean is $$\mbox{SEM} = \frac{\sigma}{ \sqrt{N} }$$ Because we divide the population standard devation $\sigma$ by the square root of the sample size $N$, the SEM gets smaller as the sample size increases. It also tells us that the shape of the sampling distribution becomes normal.

This result is useful for all sorts of things. It tells us why large experiments are more reliable than small ones, and because it gives us an explicit formula for the standard error it tells us **how much** more reliable a large experiment is. It tells us why the normal distribution is, well, **normal**. In real experiments, many of the things that we want to measure are actually averages of lots of different quantities (e.g., arguably, "general" intelligence as measured by IQ is an average of a large number of "specific" skills and abilities), and when that happens, the averaged quantity should follow a normal distribution. Because of this mathematical law, the normal distribution pops up over and over again in real data.

## Estimating population parameters [sec:pointestimates]

In all the IQ examples in the previous sections, we actually knew the population parameters ahead of time. As every undergraduate gets taught in their very first lecture on the measurement of intelligence, IQ scores are **defined** to have mean 100 and standard deviation 15. However, this is a bit of a lie. How do we know that IQ scores have a true population mean of 100? Well, we know this because the people who designed the tests have administered them to very large samples, and have then "rigged" the scoring rules so that their sample has mean 100. That's not a bad thing of course: it's an important part of designing a psychological measurement. However, it's important to keep in mind that this theoretical mean of 100 only attaches to the population that the test designers used to design the tests. Good test designers will actually go to some lengths to provide "test norms" that can apply to lots of different populations (e.g., different age groups, nationalities etc).

This is very handy, but of course almost every research project of interest involves looking at a different population of people to those used in the test norms. For instance, suppose you wanted to measure the effect of low level lead poisoning on cognitive functioning in Port Pirie, a South Australian industrial town with a lead smelter. Perhaps you decide that you want to compare IQ scores among people in Port Pirie to a comparable sample in Whyalla, a South Australian industrial town with a steel refinery. Regardless of which town you're thinking about, it doesn't make a lot of sense simply to **assume** that the true population mean IQ is 100. No-one has, to my knowledge, produced sensible norming data that can automatically be applied to South Australian industrial towns. We're going to have to **estimate** the population parameters from a sample of data. So how do we do this?

Suppose we go to Port Pirie and 100 of the locals are kind enough to sit through an IQ test. The average IQ score among these people turns out to be $\bar{X}=98.5$. So what is the true mean IQ for the entire population of Port Pirie? Obviously, we don't know the answer to that question. It could be $97.2$, but if could also be $103.5$. Our sampling isn't exhaustive so we cannot give a definitive answer. Nevertheless if I was forced at gunpoint to give a "best guess" I'd have to say $98.5$. That's the essence of statistical estimation: giving a best guess.

In this example, estimating the unknown poulation parameter is straightforward. I calculate the sample mean, and I use that as my **estimate of the population mean**. It's pretty simple, and in the next section I'll explain the statistical justification for this intuitive answer. However, for the moment what I want to do is make sure you recognise that the sample statistic and the estimate of the population parameter are conceptually different things. A sample statistic is a description of your data, whereas the estimate is a guess about the population. With that in mind, statisticians often different notation to refer to them. For instance, if true population mean is denoted $\mu$, then we would use $\hat\mu$ to refer to our estimate of the population mean. In contrast, the sample mean is denoted $\bar{X}$ or sometimes $m$. However, in simple random samples, the estimate of the population mean is identical to the sample mean: if I observe a sample mean of $\bar{X} = 98.5$, then my estimate of the population mean is also $\hat\mu = 98.5$. To help keep the notation clear, here's a handy table:

cll Symbol & What is it? & Do we know what it is?\
\
$\bar{X}$ & Sample mean & Yes, calculated from the raw data\
$\mu$ & True population mean & Almost never known for sure\
$\hat{\mu}$ & Estimate of the population mean & Yes, identical to the sample mean

So far, estimation seems pretty simple, and you might be wondering why I forced you to read through all that stuff about sampling theory. In the case of the mean, our estimate of the population parameter (i.e. $\hat\mu$) turned out to identical to the corresponding sample statistic (i.e. $\bar{X}$). However, that's not always true. To see this, let's have a think about how to construct an **estimate of the population standard deviation**, which we'll denote $\hat\sigma$. What shall we use as our estimate in this case? Your first thought might be that we could do the same thing we did when estimating the mean, and just use the sample statistic as our estimate. That's almost the right thing to do, but not quite.

Here's why. Suppose I have a sample that contains a single observation. For this example, it helps to consider a sample where you have no intutions at all about what the true population values might be, so let's use something completely fictitious. Suppose the observation in question measures the **cromulence** of my shoes. It turns out that my shoes have a cromulence of 20. So here's my sample:

`20`

This is a perfectly legitimate sample, even if it does have a sample size of $N=1$. It has a sample mean of 20, and because every observation in this sample is equal to the sample mean (obviously!) it has a sample standard deviation of 0. As a description of the **sample** this seems quite right: the sample contains a single observation and therefore there is no variation observed within the sample. A sample standard deviation of $s = 0$ is the right answer here. But as an estimate of the **population** standard deviation, it feels completely insane, right? Admittedly, you and I don't know anything at all about what "cromulence" is, but we know something about data: the only reason that we don't see any variability in the **sample** is that the sample is too small to display any variation! So, if you have a sample size of $N=1$, it **feels** like the right answer is just to say "no idea at all".

Notice that you **don't** have the same intuition when it comes to the sample mean and the population mean. If forced to make a best guess about the population mean, it doesn't feel completely insane to guess that the population mean is 20. Sure, you probably wouldn't feel very confident in that guess, because you have only the one observation to work with, but it's still the best guess you can make.

Let's extend this example a little. Suppose I now make a second observation. My data set now has $N=2$ observations of the cromulence of shoes, and the complete sample now looks like this:

`20, 22`

This time around, our sample is **just** large enough for us to be able to observe some variability: two observations is the bare minimum number needed for any variability to be observed! For our new data set, the sample mean is $\bar{X}=21$, and the sample standard deviation is $s=1$. What intuitions do we have about the population? Again, as far as the population mean goes, the best guess we can possibly make is the sample mean: if forced to guess, we'd probably guess that the population mean cromulence is 21. What about the standard deviation? This is a little more complicated. The sample standard deviation is only based on two observations, and if you're at all like me you probably have the intuition that, with only two observations, we haven't given the population "enough of a chance" to reveal its true variability to us. It's not just that we suspect that the estimate is **wrong**: after all, with only two observations we expect it to be wrong to some degree. The worry is that the error is **systematic**. Specifically, we suspect that the sample standard deviation is likely to be smaller than the population standard deviation.

[fig:sampdistsd]

This intuition feels right, but it would be nice to demonstrate this somehow. There are in fact mathematical proofs that confirm this intuition, but unless you have the right mathematical background they don't help very much. Instead, what I'll do is use  to simulate the results of some experiments. With that in mind, let's return to our IQ studies. Suppose the true population mean IQ is 100 and the standard deviation is 15. I can use the **rnorm()** function to generate the the results of an experiment in which I measure $N=2$ IQ scores, and calculate the sample standard deviation. If I do this over and over again, and plot a histogram of these sample standard deviations, what I have is the **sampling distribution of the standard deviation**. I've plotted this distribution in Figure [fig:sampdistsd]. Even though the true population standard deviation is 15, the average of the **sample** standard deviations is only 8.5. Notice that this is a very different result to what we found in Figure [fig:IQsamp]b when we plotted the sampling distribution of the mean. If you look at that sampling distribution, what you see is that the population mean is 100, and the average of the sample means is also 100.

Now let's extend the simulation. Instead of restricting ourselves to the situation where we have a sample size of $N=2$, let's repeat the exercise for sample sizes from 1 to 10. If we plot the average sample mean and average sample standard deviation as a function of sample size, you get the results shown in Figure [fig:estimatorbias]. On the left hand side (panel a), I've plotted the average sample mean and on the right hand side (panel b), I've plotted the average standard deviation. The two plots are quite different: **on average**, the average sample mean is equal to the population mean. It is an **unbiased estimator**, which is essentially the reason why your best estimate for the population mean is the sample mean. The plot on the right is quite different: on average, the sample standard deviation $s$ is **smaller** than the population standard deviation $\sigma$. It is a **biased estimator**. In other words, if we want to make a "best guess" $\hat\sigma$ about the value of the population standard deviation $\sigma$, we should make sure our guess is a little bit larger than the sample standard deviation $s$.

  ----- -----
        
   (a)   (b)
  ----- -----

[fig:estimatorbias]

The fix to this systematic bias turns out to be very simple. Here's how it works. Before tackling the standard deviation, let's look at the variance. If you recall from Section [sec:var], the sample variance is defined to be the average of the squared deviations from the sample mean. That is: $$s^2 = \frac{1}{N} \sum_{i=1}^N (X_i - \bar{X})^2$$ The sample variance $s^2$ is a biased estimator of the population variance $\sigma^2$. But as it turns out, we only need to make a tiny tweak to transform this into an unbiased estimator. All we have to do is divide by $N-1$ rather than by $N$. If we do that, we obtain the following formula: $$\hat\sigma^2 = \frac{1}{N-1} \sum_{i=1}^N (X_i - \bar{X})^2$$ This is an unbiased estimator of the population variance $\sigma$. Moreover, this finally answers the question we raised in Section [sec:var]. Why did  give us slightly different answers when we used the **var()** function? Because the **var()** function calculates $\hat\sigma^2$ not $s^2$, that's why. A similar story applies for the standard deviation. If we divide by $N-1$ rather than $N$, our estimate of the population standard deviation becomes: $$\hat\sigma = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (X_i - \bar{X})^2}$$ and when we use 's built in standard deviation function **sd()**, what it's doing is calculating $\hat\sigma$, not $s$.

One final point: in practice, a lot of people tend to refer to $\hat{\sigma}$ (i.e., the formula where we divide by $N-1$) as the **sample** standard deviation. Technically, this is incorrect: the **sample** standard deviation should be equal to $s$ (i.e., the formula where we divide by $N$). These aren't the same thing, either conceptually or numerically. One is a property of the sample, the other is an estimated characteristic of the population. However, in almost every real life application, what we actually care about is the estimate of the population parameter, and so people always report $\hat\sigma$ rather than $s$. This is the right number to report, of course, it's that people tend to get a little bit imprecise about terminology when they write it up, because "sample standard deviation" is shorter than "estimated population standard deviation". It's no big deal, and in practice I do the same thing everyone else does. Nevertheless, I think it's important to keep the two **concepts** separate: it's never a good idea to confuse "known properties of your sample" with "guesses about the population from which it came". The moment you start thinking that $s$ and $\hat\sigma$ are the same thing, you start doing exactly that.

To finish this section off, here's another couple of tables to help keep things clear:

cll Symbol & What is it? & Do we know what it is?\
\
$s$ & Sample standard deviation & Yes, calculated from the raw data\
$\sigma$ & Population standard deviation & Almost never known for sure\
$\hat{\sigma}$ & Estimate of the population & Yes, but not the same as the\
& standard deviation & sample standard deviation

cll Symbol & What is it? & Do we know what it is?\
\
$s^2$ & Sample variance & Yes, calculated from the raw data\
$\sigma^2$ & Population variance & Almost never known for sure\
$\hat{\sigma}^2$ & Estimate of the population & Yes, but not the same as the\
& variance & sample variance

## Estimating a confidence interval[sec:ci]

> **Statistics means never having to say you're certain**\
>
> – Unknown origin

Up to this point in this chapter, I've outlined the basics of sampling theory which statisticians rely on to make guesses about population parameters on the basis of a sample of data. As this discussion illustrates, one of the reasons we need all this sampling theory is that every data set leaves us with a some of uncertainty, so our estimates are never going to be perfectly accurate. The thing that has been missing from this discussion is an attempt to **quantify** the amount of uncertainty that attaches to our estimate. It's not enough to be able guess that, say, the mean IQ of undergraduate psychology students is 115 (yes, I just made that number up). We also want to be able to say something that expresses the degree of certainty that we have in our guess. For example, it would be nice to be able to say that there is a 95% chance that the true mean lies between 109 and 121. The name for this is a **confidence interval** for the mean.

Armed with an understanding of sampling distributions, constructing a confidence interval for the mean is actually pretty easy. Here's how it works. Suppose the true population mean is $\mu$ and the standard deviation is $\sigma$. I've just finished running my study that has $N$ participants, and the mean IQ among those participants is $\bar{X}$. We know from our discussion of the central limit theorem (Section [sec:clt]) that the sampling distribution of the mean is approximately normal. We also know from our discussion of the normal distribution Section [sec:normal] that there is a 95% chance that a normally-distributed quantity will fall within two standard deviations of the true mean. To be more precise, we can use the **qnorm()** function to compute the 2.5th and 97.5th percentiles of the normal distribution

\> @usrqnorm( p = c(.025, .975) ) [1] -1.959964 1.959964

Okay, so I lied earlier on. The more correct answer is that 95% chance that a normally-distributed quantity will fall within 1.96 standard deviations of the true mean. Next, recall that the standard deviation of the sampling distribution is referred to as the standard error, and the standard error of the mean is written as SEM. When we put all these pieces together, we learn that there is a 95% probability that the sample mean $\bar{X}$ that we have actually observed lies within 1.96 standard errors of the population mean. Mathematically, we write this as: $$\mu - \left( 1.96 \times \mbox{SEM} \right) \ \leq \  \bar{X}\  \leq \  \mu + \left( 1.96 \times \mbox{SEM} \right)$$ where the SEM is equal to $\sigma / \sqrt{N}$, and we can be 95% confident that this is true. However, that's not answering the question that we're actually interested in. The equation above tells us what we should expect about the sample mean, given that we know what the population parameters are. What we **want** is to have this work the other way around: we want to know what we should believe about the population parameters, given that we have observed a particular sample. However, it's not too difficult to do this. Using a little high school algebra, a sneaky way to rewrite our equation is like this: $$\bar{X} -  \left( 1.96 \times \mbox{SEM} \right) \ \leq \ \mu  \ \leq  \ \bar{X} +  \left( 1.96 \times \mbox{SEM}\right)$$ What this is telling is is that the range of values has a 95% probability of containing the population mean $\mu$. We refer to this range as a **95% confidence interval**, denoted $\mbox{CI}_{95}$. In short, as long as $N$ is sufficiently large – large enough for us to believe that the sampling distribution of the mean is normal – then we can write this as our formula for the 95% confidence interval: $$\mbox{CI}_{95} = \bar{X} \pm \left( 1.96 \times \frac{\sigma}{\sqrt{N}} \right)$$ Of course, there's nothing special about the number 1.96: it just happens to be the multiplier you need to use if you want a 95% confidence interval. If I'd wanted a 70% confidence interval, I could have used the **qnorm()** function to calculate the 15th and 85th quantiles:

\> @usrqnorm( p = c(.15, .85) ) [1] -1.036433 1.036433

and so the formula for $\mbox{CI}_{70}$ would be the same as the formula for $\mbox{CI}_{95}$ except that we'd use 1.04 as our magic number rather than 1.96.

As usual, I lied. The formula that I've given above for the 95% confidence interval is approximately correct, but I glossed over an important detail in the discussion. Notice my formula requires you to use the standard error of the mean, SEM, which in turn requires you to use the true population standard deviation $\sigma$. Yet, in Section [sec:pointestimates] I stressed the fact that we don't actually **know** the true population parameters. Because we don't know the true value of $\sigma$, we have to use an estimate of the population standard deviation $\hat{\sigma}$ instead. This is pretty straightforward to do, but this has the consequence that we need to use the quantiles of the $t$-distribution rather than the normal distribution to calculate our magic number; and the answer depends on the sample size. When $N$ is very large, we get pretty much the same value using **qt()** that we would if we used **qnorm()**...

\> @usrN \<- 10000 \# suppose our sample size is 10,000 \> @usrqt( p = .975, df = N-1) \# calculate the 97.5th quantile of the t-dist [1] 1.960201

But when $N$ is small, we get a much bigger number when we use the $t$ distribution:

\> @usrN \<- 10 \# suppose our sample size is 10 \> @usrqt( p = .975, df = N-1) \# calculate the 97.5th quantile of the t-dist [1] 2.262157

There's nothing too mysterious about what's happening here. Bigger values mean that the confidence interval is wider, indicating that we're more uncertain about what the true value of $\mu$ actually is. When we use the $t$ distribution instead of the normal distribution, we get bigger numbers, indicating that we have more uncertainty. And why do we have that extra uncertainty? Well, because our estimate of the population standard deviation $\hat\sigma$ might be wrong! If it's wrong, it implies that we're a bit less sure about what our sampling distribution of the mean actually looks like... and this uncertainty ends up getting reflected in a wider confidence interval.

The hardest thing about confidence intervals is understanding what they **mean**. Whenever people first encounter confidence intervals, the first instinct is almost always to say that "there is a 95% probabaility that the true mean lies inside the confidence interval". It's simple, and it seems to capture the common sense idea of what it means to say that I am "95% confident". Unfortunately, it's not quite right. The intuitive definition relies very heavily on your own personal **beliefs** about the value of the population mean. I say that I am 95% confident because those are my beliefs. In everyday life that's perfectly okay, but if you remember back to Section [sec:probmeaning], you'll notice that talking about personal belief and confidence is a Bayesian idea. Personally (speaking as a Bayesian) I have no problem with the idea that the phrase "95% probability" is allowed to refer to a personal belief. However, confidence intervals are **not** Bayesian tools. Like everything else in this chapter, confidence intervals are **frequentist** tools, and if you are going to use frequentist methods then it's not appropriate to attach a Bayesian interpretation to them. If you use frequentist methods, you must adopt frequentist interpretations!

Okay, so if that's not the right answer, what is? Remember what we said about frequentist probability: the only way we are allowed to make "probability statements" is to talk about a sequence of events, and to count up the frequencies of different kinds of events. From that perspective, the interpretation of a 95% confidence interval must have something to do with replication. Specifically: if we replicated the experiment over and over again and computed a 95% confidence interval for each replication, then 95% of those **intervals** would contain the true mean. More generally, 95% of all confidence intervals constructed using this procedure should contain the true population mean. This idea is illustrated in Figure [fig:cirep], which shows 50 confidence intervals constructed for a "measure 10 IQ scores" experiment (top panel) and another 50 confidence intervals for a "measure 25 IQ scores" experiment (bottom panel). A bit fortuitously, across the 100 replications that I simulated, it turned out that exactly 95 of them contained the true mean.

cc &\
\
 &

[fig:cirep]

The critical difference here is that the Bayesian claim makes a probability statement about the population mean (i.e., it refers to our uncertainty about the population mean), which is not allowed under the frequentist interpretation of probability because you can't "replicate" a population! In the frequentist claim, the population mean is fixed and no probabilistic claims can be made about it. Confidence intervals, however, are repeatable so we can replicate experiments. Therefore a frequentist is allowed to talk about the probability that the **confidence interval** (a random variable) contains the true mean; but is not allowed to talk about the probability that the **true population mean** (not a repeatable event) falls within the confidence interval.

I know that this seems a little pedantic, but it does matter. It matters because the difference in interpretation leads to a difference in the mathematics. There is a Bayesian alternative to confidence intervals, known as **credible intervals**. In most situations credible intervals are quite similar to confidence intervals, but in other cases they are drastically different. As promised, though, I'll talk more about the Bayesian perspective in Chapter [ch:bayes].

As far as I can tell, the core packages in  don't include a simple function for calculating confidence intervals for the mean. They **do** include a lot of complicated, extremely powerful functions that can be used to calculate confidence intervals associated with lots of different things, such as the **confint()** function that we'll use in Chapter [ch:regression]. But I figure that when you're first learning statistics, it might be useful to start with something simpler. As a consequence, the **lsr** package includes a function called **ciMean()** which you can use to calculate your confidence intervals. There are two arguments that you might want to specify:

Hopefully that's fairly clear.

[fig:bargraph.CI]

[fig:lineplot.CI]

[fig:plotmeans]

There's several different ways you can draw graphs that show confidence intervals as error bars. I'll show three versions here, but this certainly doesn't exhaust the possibilities. In doing so, what I'm assuming is that you want to draw is a plot showing the means and confidence intervals for one variable, broken down by different levels of a second variable. For instance, in our **afl** data that we discussed earlier, we might be interested in plotting the average **attendance** by **year**. I'll do this using three different functions, **bargraph.CI()**, **lineplot.CI()** (both of which are in the **sciplot** package), and **plotmeans()** (which is in the **gplots**) package. Assuming that you've installed these packages on your system (see Section [sec:packageinstall] if you've forgotten how to do this), you'll need to load them. You'll also need to load the **lsr** package, because we'll make use of the **ciMean()** function to actually calculate the confidence intervals

\> @usrlibrary( sciplot ) \# bargraph.CI() and lineplot.CI() functions \> @usrlibrary( gplots ) \# plotmeans() function \> @usrlibrary( lsr ) \# ciMean() function

Here's how to plot the means and confidence intervals drawn using **bargraph.CI()**.

\> @usrbargraph.CI( x.factor = year, \# grouping variable + @usr response = attendance,  \# outcome variable + @usr data = afl,  \# data frame with the variables + @usr ci.fun= ciMean, \# name of the function to calculate CIs + @usr xlab = "Year",  \# x-axis label + @usr ylab = "Average Attendance" \# y-axis label + @usr)

This produces the output shown in Figure [fig:bargraph.CI]. We can use the same arguments when calling the **lineplot.CI()** function:

\> @usrlineplot.CI( x.factor = year, \# grouping variable + @usr response = attendance,  \# outcome variable + @usr data = afl,  \# data frame with the variables + @usr ci.fun= ciMean, \# name of the function to calculate CIs + @usr xlab = "Year",  \# x-axis label + @usr ylab = "Average Attendance" \# y-axis label + @usr)

And the output for this command is shown in Figure [fig:lineplot.CI]. Finally, here's how you would do it using **plotmeans()**:

\> @usrplotmeans( formula = attendance   year, \# outcome   group + @usr data = afl,  \# data frame with the variables + @usr n.label = FALSE \# don't show the sample sizes + @usr)

This is shown in Figure [fig:plotmeans].

## Summary

In this chapter I've covered two main topics. The first half of the chapter talks about sampling theory, and the second half talks about how we can use sampling theory to construct estimates of the population parameters. The section breakdown looks like this:

-   Basic ideas about samples, sampling and populations (Section [sec:srs])

-   Statistical theory of sampling: the law of large numbers (Section [sec:lawlargenumbers]), sampling distributions and the central limit theorem (Section [sec:samplesandclt]).

-   Estimating means and standard deviations (Section [sec:pointestimates])

-   Estimating a confidence interval (Section [sec:ci])

As always, there's a lot of topics related to sampling and estimation that aren't covered in this chapter, but for an introductory psychology class this is fairly comprehensive I think. For most applied researchers you won't need much more theory than this. One big question that I haven't touched on in this chapter is what you do when you don't have a simple random sample. There is a lot of statistical theory you can draw on to handle this situation, but it's well beyond the scope of this book.
