# Mixed Design ANOVA

```{r}
library(ggplot2)
```


We should probably call this chapter "more on factorial designs", rather than just mixed design ANOVA. We are going to do a couple things in this chapter. The most important thing we do is give you more exposure to factorial designs. The second thing we do is show that you can mix it up with ANOVA. You already know that you can have more than one IV. And, you know that research designs can be between-subjects or within-subjects (repeated-measures). When you have more than one IV, they can all be between-subjects variables, they can all be within-subject repeated measures, or they can be a mix: say one between-subject variable and one within-subject variable. You can use ANOVA to anlayse all of these kinds of designs. You always get one main effect for each IV, and a number of interactions, or just one, depending on the number of IVs.

## Looking at main effects and interactions

Designs with multiple factors are very common. When you read a research article you will often see graphs that show the results from designs with multiple factors. It would be good for you if you were comfortable interpreting the meaning of those results. The skill here is to be able to look at a graph and see the pattern of main effects and interactions. This skill is important, because the patterns in the data can quickly become very complicated looking, especially when there are more than two independent variables, with more than two levels.

### 2x2 designs

Let's take the case of 2x2 designs. There will always be the possibility of two main effects and one interaction. You will always be able to compare the means for each main effect and interaction. If the means are different then there is a main effect or interaction. Here's the thing, there a bunch of ways all of this can turn out. Check out the ways, there are 8 of them:

1. no IV1 main effect, no IV2 main effect, no interaction
2. IV1 main effect, no IV2 main effect, no interaction
3. IV1 main effect, no IV2 main effect, interaction
4. IV1 main effect, IV2 main effect, no interaction
5. IV1 main effect, IV2 main effect, interaction
6. no IV1 main effect, IV2 main effect, no interaction
7. no IV1 main effect, IV2 main effect, interaction
8. no IV1 main effect, no IV2 main effect, interaction

OK, so if you run a 2x2, any of these 8 general patterns could occur in your data. That's a lot to keep track of isn't. As you develop your skills in examining graphs that plot means, you should be able to look at the graph and visually guesstimate if there is, or is not,  a main effect or interaction. You will need you inferential statistics to tell you for sure, but it is worth knowing how to know see the patterns.

In this section we show you some example patterns so that you can get some practice looking at the patterns. First, in bar graph form. Note, we used the following labels for the graph:

- 1 = there was a main effect for IV1. 
- ~1 = there was not a main effect for IV1
- 2 = there was a main effect for IV2
- ~2 = there was not a main effect of IV2
- 1x2 = there was an interaction
` ~1x2 = there was not an interaction

```{r}
p1<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(5,5,5,5))

p2<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(10,10,5,5))

p3<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(10,13,5,2))

p4<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(5,10,10,15))

p5<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(10,18,5,7))

p6<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(5,10,5,10))

p7<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(2,12,5,9))

p8<- data.frame(IV1 = c("A","A","B","B"),
                IV2 = c("1","2","1","2"),
                means = c(5,10,10,5))

all_22s <- rbind(p1,p2,p3,p4,p5,p6,p7,p8)

type <- c(rep("~1, ~2, ~1x2",4),
          rep("1, ~2, ~1x2",4),
          rep("1, ~2, 1x2",4),
          rep("1, 2, ~1x2",4),
          rep("1, 2, 1x2",4),
          rep("~1, 2, ~1x2",4),
          rep("~1, 2, 1x2",4),
          rep("~1, ~2, 1x2",4))

type<-as.factor(type)

all_22s <- cbind(all_22s,type)

ggplot(all_22s, aes(x=IV1, y=means, group=IV2, fill=IV2))+
  geom_bar(stat="identity", position="dodge")+
  theme_classic()+
  facet_wrap(~type, nrow=2)+
  theme(legend.position = "bottom")
                             
```


Next, we show you the same thing in line graph form:


```{r}

ggplot(all_22s, aes(x=IV1, y=means, group=IV2, color=IV2))+
  geom_point()+
  geom_line()+
  theme_classic()+
  facet_wrap(~type, nrow=2)+
  theme(legend.position = "bottom")
                             
```

You might find the line graphs easier to interpret. Whenever the lines cross, or would cross if they kept going, you have a possibility of an interaction. Whenever the lines are parallel, there can't be an interaction. When both of the points on the A side are higher or lower than both of the points on the B side, then you have a main effect for IV1 (A vs B). Whenever the green line is above or below the red line, then you have a main effect for IV2 (1 vs. 2). We know this is complicated. You should see what all the possibilities look like when we start adding more levels or more IVs. It gets nuts. Because of this nuttiness, it is often good practice to make your research designs simple (as few IVs and levels as possible to test your question). That way it will be easier to interpret your data. Whenever you see that someone ran a 4x3x7x2 design, your head should spin. It's just too complicated.

## Interactions can make main effects even when they aren't really there



## A general model

It's probably time we introduce you to another new concept. This concept has been underyling everything we have been doing so far. It's called the general linear model, or GLM for short. Don't worry it's not that scary. And, right now, we won't dive off into the deep mind. Instead, we will do a little bit of light GLM, for beginners. The purpose is to give you some idea where the main effects and interaction come from, and when they are there, how they influence the pattern in the means.

The GLM makes you think about the various things that influence each mean you measure. It's like thinking that each mean is a little bit of this, and a little bit of that. Here's a general formula for the means:

$ \text{A mean} = \text{Grand Mean} + \text{IV1 effect} + \text{IV2 effect} + \text{Interaction effect} + \text{error} $

Let's say you ran an experiment with two IVs, but they did absolutely nothing, and there was no interaction. What would you expect your means to be in each of the four conditions. You would still measure a bunch of scores, but they should all be the same right? If so, then the grand mean of all the scores should be your mean for all of the conditions. The formula would look like this:

$ \text{Each mean} = \text{Grand Mean} + 0 + 0 + 0 + \text{error}$

Remember there will still be some sampling error, so each mean will be the grand mean, plus or minus some amount of random sampling error.






