# Eugenics, Statistics, & Psychology

This is an appendix with resources to learn more about the confluence of eugenics, statistics, and Psychology

## Websites

- [Eugenics Archives](http://eugenicsarchive.ca) This is a very comprehensive online resource about eugenics, with a focus on eugenic practices in Western Canadian provinces, but also a very well-rounded and extensive broader resource. For example, the timeline portion is very informative.
- [Cold Spring Harbor Laboratory](http://library.cshl.edu/special-collections/eugenics) The national American Eugenic Records Office (ERO) was located in what is now Cold Springs Harbor Laboratory, in Long Island, New York.

## Journal Articles

```{r, results='asis', echo=FALSE}
library(knitcitations)
library(RefManageR)
bib <- read.bibtex("eugenic.bib")

in_text_ref <- function(x){
  b<-capture.output(print(x,.opts = list(bib.style = "authoryear", first.inits = TRUE, style = "markdown",no.print.fields=c("URL","urldate","issn","doi"))))
  cat("- ")
  cat(b,collapse=" ")
  
}
```



```{r, results='asis', echo=FALSE}
in_text_ref(bib["galtonHereditaryTalentCharacter1865"])
```
 - This is where eugenics begins in the writing of Sir Francis Galton. The original is online


## Books

```{r, results='asis', echo=FALSE}
in_text_ref(bib["pressEugenicMindProject"])
```

```{r, results='asis', echo=FALSE}
in_text_ref(bib["kaminSciencePoliticsIQ1974"])
```

```{r, results='asis', echo=FALSE}
in_text_ref(bib["bashfordOxfordHandbookHistory2010"])
```


